---
title: "Week 4 Exercises"
author: "J.Blondeau"
date: "2024-09-12"
output: 
  html_document:
    code_folding: show
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
options(width = 750)
```

Import libraries but don't show the output 'include=FALSE'
```{r results='hide', message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
```

This week we will continue wrangling data in the Tidyverse.  The following are some very common functions you will use in your everyday data world.

- join()
- mutate()
- case_when()
- group_by()
- summarize()


All the exercises for this week use the "PSU_Fish_Density.csv" and the "taxonomic_data.csv" data sets.

1. Add the common name (only common name) of each species in the density data set as a new field.
    - Reorder the columns to be more readable.
  
```{r}
data <- read_csv("../data/PSU_Fish_Density.csv")
taxo <- read_csv("../data/taxonomic_data.csv")

data <- data %>%
  left_join(select(taxo, c("SPECIES_CD", "COMNAME")), by = "SPECIES_CD") %>% 
  select(YEAR:PRIMARY_SAMPLE_UNIT, COMNAME, SPECIES_CD, density)
  
print.data.frame(head(data))
```
  

2. Add a new field to the density data set named "fishery_target" at set all to TRUE

```{r}
data <- data %>% 
  mutate(fishery_target = TRUE)

print.data.frame(head(data))
```


3. Add a new field named "group" at set the value based on fish species where...
    - grouper = Coney, Red hind, 
    - sanpper = Mutton, Gray, Yellowtail
    - parrotfish = Stoplight
    - other = Triggerfish, Hogfish
<br>    

```{r}

data <- data %>%
  mutate(group = case_when(
    COMNAME %in% c("coney", "red hind")                                    ~ "grouper",
    COMNAME %in% c("mutton snapper", "gray snapper", "yellowtail snapper") ~ "snapper",
    COMNAME %in% c("stoplight parrotfish")                                 ~ "parrotfish",
    COMNAME %in% c("queen triggerfish", "hogfish")                         ~ "other",
    TRUE                                                                   ~ NA_character_
  ))

print.data.frame(head(data))
```

4. Using the group_by function, how many unique PRIMARY_SAMPLE_UNITS were sampled in each YEAR?

```{r}
data %>% 
  group_by(YEAR) %>% 
  summarise(total = n_distinct(PRIMARY_SAMPLE_UNIT)) %>% 
  ungroup() %>% 
  print.data.frame()
```


5. How many unique PRIMARY_SAMPLE_UNITS were sampled in each YEAR and PROT combination?

```{r}
data %>% 
  group_by(YEAR, PROT) %>% 
  summarise(total = n_distinct(PRIMARY_SAMPLE_UNIT)) %>% 
  ungroup() %>% 
  print.data.frame()
```

6. How many unique PRIMARY_SAMPLE_UNITS were sampled in each YEAR, PROT and STRAT combination?

```{r}
data %>% 
  group_by(YEAR, PROT, STRAT) %>% 
  summarise(total = n_distinct(PRIMARY_SAMPLE_UNIT)) %>% 
  ungroup() %>% 
  print.data.frame()
```

7. What is the difference between the following two expressions?  The summarize and mutate calls after the group_by do very different things.  In what situations would you use them?\
    data here is the "PSU_Fish_Density" data set
    - data %>% 
        <br></br>
        group_by(YEAR) %>% 
        <br></br>
        summarize( n = n_distinct(PRIMARY_SAMPLE_UNIT))
```{r}
data %>% 
  group_by(YEAR) %>% 
  summarize(n = n_distinct(PRIMARY_SAMPLE_UNIT)) %>% 
  ungroup() %>% 
  print.data.frame()
```


    - data %>%
        <br></br>
        group_by(YEAR) %>% 
        <br></br>
        mutate( n = n_distinct(PRIMARY_SAMPLE_UNIT))
        
```{r}
data %>% 
  group_by(YEAR) %>% 
  mutate(n = n_distinct(PRIMARY_SAMPLE_UNIT)) %>% 
  select(YEAR, PRIMARY_SAMPLE_UNIT, n) %>% 
  ungroup() %>% 
  print(n = 5)

```
        
        
8. Create a new dataframe that shows mean density of each species per year...hint
    - YEAR  | SPECIES_CD  | meanDensity
      ----- | ----------  | -----------
      2017  | BAL VETU    |    0.224   
      2019  | CEP FULV    |    0.685
      

```{r}
spp_dens_yr <- data %>% 
  group_by(YEAR, SPECIES_CD) %>% 
  summarise(meanDensity = mean(density)) %>% 
  ungroup()

print.data.frame(spp_dens_yr)
```


9. Create a new dataframe that shows mean density of each species in each PROT per year...hint
    - YEAR  | SPECIES_CD  | PROT | meanDensity
      ----- | ----------  | -----|-----------
      2017  | BAL VETU    | 0    |0.229  
      2019  | CEP FULV    | 1    |0.308

```{r}
spp_dens_prot_yr <- data %>% 
  group_by(YEAR, PROT, SPECIES_CD) %>% 
  summarise(meanDensity = mean(density)) %>% 
  ungroup()

print.data.frame(spp_dens_prot_yr)
```

10. In the "PSU_Fish_Density.csv" data set, the PROT field refers to sites that are inside the VI National Park (PROT = 1) and sites that are outside the Park (PROT = 0).  How many of the 4 groups of species had higher densities inside the National Park?

```{r}
group_dens_prot <- data %>% 
  group_by(PROT, group) %>% 
  summarise(group_dens = mean(density)) %>% 
  ungroup() %>% 
  arrange(group, PROT)

print.data.frame(group_dens_prot)
```

